---
title: "Demonstrating Global Data Visualization Techniques with R and Plotly"
author: "Liz Kernan"
date: "2025-12-16"
categories: [R, visualizations]
format:
  html:
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = TRUE)
```

### Intro

This project showcases how publicly available global datasets can be combined with modern visualization tools in R to explore broad patterns in social behavior and economic indicators. Using the Gallup World Poll’s question *“Have you helped a stranger in the past month?”* and ten years of GDP per capita data (2015–2024), the goal is not to make causal claims—but to demonstrate techniques for preparing multi-year datasets and building interactive, exploratory graphics with `plotly`, `ggplot2`, and `gganimate`.

The workflow moves from static summaries to global choropleth maps, animated scatterplots, and simple modeling. By showing how each step builds on the last, the example illustrates how visualization can help reveal patterns that might otherwise be lost in raw tables or static snapshots.

```{r message = FALSE,warning=FALSE}
#load packages 
library(dplyr)
library(tidyverse)
library(plotly)
library(crosstalk)
library(DT)
library(ggplot2)
library(gganimate)
library(plotrix)
library(gapminder)
library(viridis)
library(countrycode)
library(ggrepel)
```

```{r, message=FALSE, warning=FALSE}
#load datasets and join

setwd("X:/rcs_admin_sharedspace/ekernan/happiness_and_taxes/")
countryISO3 <- read.csv("countrycode.csv")
gdp <- read.csv("gdp_per_capita_2015_2024.csv")
helped_stranger <- read.csv("gallup_helped_stranger.csv")
helped_stranger <- helped_stranger[-c(1383:1393), ]

# Convert to ISO3 country codes
helped_stranger$ISO <- countrycode(helped_stranger$country, origin = "country.name", destination = "iso3c")
helped_stranger <- helped_stranger %>% select(-country)

gdp <- gdp %>%
  pivot_longer(
    cols = starts_with("X"),      # pick all columns starting with "X"
    names_to = "year",
    values_to = "value"
  ) %>%
  mutate(year = as.integer(sub("X", "", year)))   # remove "X" and convert to number

gdp_avg <- gdp %>%
  group_by(ISO) %>%
  summarise(avg_gdp = mean(value, na.rm = TRUE))

help_avg <- helped_stranger %>%
  group_by(ISO) %>%
  summarise(avg_help = mean(helped_a_stranger, na.rm = TRUE))

gdp_helped_stranger <- merge(helped_stranger,gdp, by=c("ISO","year"))
gdp_helped_stranger <- merge(gdp_helped_stranger, gdp_avg, by = "ISO")
gdp_helped_stranger <- merge(gdp_helped_stranger, help_avg, by = "ISO")
gdp_helped_stranger$helped_a_stranger<- gdp_helped_stranger$helped_a_stranger * 100
gdp_helped_stranger$avg_help<- gdp_helped_stranger$avg_help * 100

write.csv(gdp_helped_stranger, "pd_data.csv", row.names = FALSE)

avg_help_by_year <- gdp_helped_stranger %>%
  group_by(year) %>%
  summarise(avg_helped = mean(helped_a_stranger, na.rm = TRUE))
```

### Average percentage of willingness to help a stranger over time (2015-2024)

```{r, warning=FALSE, message=FALSE, fig.align='center', out.width='70%'}
ggplot(avg_help_by_year, aes(x = factor(year), y = avg_helped,
                             fill = year == 2020)) +   # conditional fill
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "steelblue"), guide = FALSE) +  # red for 2020, blue for others
  geom_text(
    aes(label = paste0(round(avg_helped), "%")),
    vjust = -0.5,
    size = 4
  ) +
  labs(
    title = "Average Helped a Stranger by Year",
    x = "Year",
    y = "Average Helped a Stranger (%)"
  ) +
  theme_minimal() +
  ylim(0, max(avg_help_by_year$avg_helped) * 1.1)
```

A basic bar chart provides a first look at changes over time in the percentage of people who reported helping a stranger. Highlighting 2020 in red draws attention to the onset of the COVID-19 pandemic, a natural point of comparison when examining shifts in prosocial behavior. While simple, this plot demonstrates the value of computed summaries—here, yearly averages across all available countries.

### Global choropleth map: Plotly for interactive geospatial visualization

```{r, message=FALSE, warning=FALSE}

plot_ly(
  gdp_helped_stranger,
  type = "choropleth",
  locations = ~ISO,
  z = ~helped_a_stranger * 1,
  frame = ~year,
  colorscale = "Reds",
  text = ~paste(
    country),
  colorbar = list(
    title = "% Helped a Stranger",
    tickformat = ".1f"
  )
) %>%
  layout(
    title = "Percentage of willingness to help a stranger per country 2015-2024",
    geo = list(
      projection = list(type = "natural earth"),  # Natural Earth projection
      showframe = TRUE,                           # <-- draws the outer border
      framecolor = "black",                       # <-- color of the outer frame
      framewidth = 2,                             # <-- thickness of the outer frame
      showcoastlines = TRUE,
      coastlinecolor = "gray50",
      showocean = TRUE,                           # <-- adds ocean fill
      oceancolor = "lightblue",                   # <-- ocean color
      landcolor = "lightgray",
      showlakes = TRUE,
      lakecolor = "white",
      showcountries = TRUE,
      countrycolor = "gray60"
    )
  ) %>%
  animation_opts(
    frame = 800,
    transition = 0,
    redraw = TRUE
  )

```

To explore geographic variation in helping behavior, the first interactive visualization uses a Plotly choropleth map. Each frame of the animation corresponds to a single year from 2015 to 2024, allowing users to drag the slider or press “play” to watch global patterns evolve.

Because the map is interactive, users can:

-   Hover over any country to view its percentage

-   Rotate and zoom to inspect regions

-   Compare year-to-year differences with the animation slider

-   Examine areas of consistently high or low prosociality

This visualization emphasizes how interactivity helps users move beyond static summaries—particularly when patterns vary widely across continents.

#### \^\^code used

``` r
plot_ly(
  gdp_helped_stranger,
  type = "choropleth",
  locations = ~ISO,
  z = ~helped_a_stranger * 1,
  frame = ~year,
  colorscale = "Reds",
  text = ~paste(country),
  colorbar = list(
    title = "% Helped a Stranger",
    tickformat = ".1f"
  )
) %>%
  layout(
    title = "Percentage of willingness to help a stranger per country 2015-2024",
    geo = list(
      projection = list(type = "natural earth"),
      showframe = TRUE,
      framecolor = "black",
      framewidth = 2,
      showcoastlines = TRUE,
      coastlinecolor = "gray50",
      showocean = TRUE,
      oceancolor = "lightblue",
      landcolor = "lightgray",
      showlakes = TRUE,
      lakecolor = "white",
      showcountries = TRUE,
      countrycolor = "gray60"
    )
  ) %>%
  animation_opts(
    frame = 800,
    transition = 0,
    redraw = TRUE
  )
```

### Comparing highest and lowest GDP countries (2015 vs. 2024)

```{r, message=FALSE, warning=FALSE}
# helper function to produce the plot for any year
plot_top_bottom_by_year <- function(data, year_val) {
  # select year and top/bottom 10
  df_year <- data %>% filter(year == year_val)
  
  df_top10 <- df_year %>%
    arrange(desc(value)) %>%
    slice_head(n = 10) %>%
    mutate(group = "Top 10 GDP")
  
  df_bottom10 <- df_year %>%
    arrange(value) %>%
    slice_head(n = 10) %>%
    mutate(group = "Bottom 10 GDP")
  
  df_compare <- bind_rows(df_top10, df_bottom10) %>%
    # convert help to percent and build formatted label
    mutate(
      help_percent = helped_a_stranger * 1,
      label_text = paste0(country, " (", sprintf("%.2f%%", help_percent), ")")
    )
  
  ggplot(df_compare, aes(x = value, y = help_percent, color = group, label = label_text)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3, max.overlaps = Inf, segment.color = "grey60") +
    scale_color_manual(values = c("Top 10 GDP" = "blue", "Bottom 10 GDP" = "red")) +
    scale_x_continuous(labels = scales::comma) +
    labs(
      x = paste0("GDP per capita (", year_val, ")"),
      y = "% who helped a stranger",
      color = "Group",
      title = paste0("Top 10 vs Bottom 10 GDP Countries in ", year_val)
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "top"
    )
}

# Example usage
plot_top_bottom_by_year(gdp_helped_stranger, 2015)
plot_top_bottom_by_year(gdp_helped_stranger, 2024)
```

To explore the relationship between income and prosocial behaviors, we can compare the countries with the highest and lowest GDP per capita in selected years. A helper function extracts the top and bottom ten economies and plots them against the percentage of people who helped a stranger. This produces a scatterplot that makes it easy to see whether wealthier or poorer nations tend to differ in reported helping behavior.

Repeating the same function for the first and the last years of the dataset shows how the grouping and relationships shift over time. Using the same function to display the 2020 plot offers a pandemic-year snapshot without requiring new code.

### Visualizing the pandemic year (2020)

```{r, message=FALSE, warning=FALSE}
plot_top_bottom_by_year(gdp_helped_stranger, 2020)
```

### Animated comparison of top and bottom GDP groups over time (Plotly)

```{r, message=FALSE, warning=FALSE}

# prepare per-year top10 & bottom10
gdp_filtered <- gdp_helped_stranger %>%
  group_by(year) %>%
  arrange(year, desc(value)) %>%
  mutate(rank_desc = row_number(),
         n_in_year = n()) %>%
  filter(rank_desc <= 10 | rank_desc > (n_in_year - 10)) %>%
  ungroup() %>%
  mutate(
    color_group = ifelse(rank_desc <= 10, "Top 10 GDP", "Bottom 10 GDP"),
    helped_percent = helped_a_stranger * 1,                           # convert to percent
    label_text = paste0(country, " (", sprintf("%.2f%%", helped_percent), ")")
  )

# Plotly: formatting similar to your ggplot
plot_ly(
  data = gdp_filtered,
  x = ~value,
  y = ~helped_percent,
  frame = ~year,
  text = ~label_text,
  color = ~color_group,
  colors = c("Top 10 GDP" = "blue", "Bottom 10 GDP" = "red"),
  type = "scatter",
  mode = "markers+text",
  textposition = "middle right",        # label beside marker
  marker = list(size = 10, opacity = 0.85),
  textfont = list(size = 10),
  cliponaxis = FALSE                    # prevents Plotly from cropping labels at axis
) %>%
  layout(
    title = "Top 10 (blue) vs Bottom 10 (red) GDP Countries — GDP vs % Helped over Time",
    xaxis = list(
      title = "GDP per capita",
      range = c(0, max(gdp_filtered$value, na.rm = TRUE) * 1.05),
      tickformat = ",",                  # comma formatting for numbers
      automargin = TRUE
    ),
    yaxis = list(
      title = "% who helped a stranger",
      tickformat = ".2f",
      ticksuffix = "%",
      range = c(0, max(gdp_filtered$helped_percent, na.rm = TRUE) * 1.1),
      automargin = TRUE
    ),
    margin = list(l = 140, r = 40, t = 80, b = 60),   # more left margin so labels fit
    legend = list(title = list(text = "Group"))
  ) %>%
  animation_opts(
    frame = 800,
    transition = 0,
    redraw = TRUE,
    easing = "linear"
  ) %>%
  animation_slider(currentvalue = list(prefix = "Year: ")) %>%
  config(displayModeBar = TRUE)
```

Building on the static scatterplots, the second interactive Plotly visualization animates the relationship between GDP and the willingness to help strangers across the entire 10-year window. For each year, the top and bottom 10 economies are identified and plotted, with color distinguishing high- and low-GDP country groups.

This animated scatterplot allows users to:

-   Follow specific countries across time

-   Watch how high-GDP and low-GDP groups separate or converge

-   Hover to reveal country names and exact percentages

-   Observe whether prosocial behavior shifts with economic trends

Because the animation updates yearly, transitions highlight movement that might be less apparent when looking at isolated years. This provides a more intuitive sense of economic cycles and how social indicators might respond.

#### \^\^code used

``` r
# prepare per-year top10 & bottom10
gdp_filtered <- gdp_helped_stranger %>%
  group_by(year) %>%
  arrange(year, desc(value)) %>%
  mutate(
    rank_desc = row_number(),
    n_in_year = n()
  ) %>%
  filter(rank_desc <= 10 | rank_desc > (n_in_year - 10)) %>%
  ungroup() %>%
  mutate(
    color_group = ifelse(rank_desc <= 10, "Top 10 GDP", "Bottom 10 GDP"),
    helped_percent = helped_a_stranger * 1,
    label_text = paste0(country, " (", sprintf("%.2f%%", helped_percent), ")")
  )

# Plotly: formatting similar to your ggplot
plot_ly(
  data = gdp_filtered,
  x = ~value,
  y = ~helped_percent,
  frame = ~year,
  text = ~label_text,
  color = ~color_group,
  colors = c("Top 10 GDP" = "blue", "Bottom 10 GDP" = "red"),
  type = "scatter",
  mode = "markers+text",
  textposition = "middle right",
  marker = list(size = 10, opacity = 0.85),
  textfont = list(size = 10),
  cliponaxis = FALSE
) %>%
  layout(
    title = "Top 10 (blue) vs Bottom 10 (red) GDP Countries — GDP vs % Helped over Time",
    xaxis = list(
      title = "GDP per capita",
      range = c(0, max(gdp_filtered$value, na.rm = TRUE) * 1.05),
      tickformat = ",",
      automargin = TRUE
    ),
    yaxis = list(
      title = "% who helped a stranger",
      tickformat = ".2f",
      ticksuffix = "%",
      range = c(0, max(gdp_filtered$helped_percent, na.rm = TRUE) * 1.1),
      automargin = TRUE
    ),
    margin = list(l = 140, r = 40, t = 80, b = 60),
    legend = list(title = list(text = "Group"))
  ) %>%
  animation_opts(
    frame = 800,
    transition = 0,
    redraw = TRUE,
    easing = "linear"
  ) %>%
  animation_slider(currentvalue = list(prefix = "Year: ")) %>%
  config(displayModeBar = TRUE)
```

### Statistical modeling as a complement to visualization

```{r, warning=FALSE, message=FALSE}
model_scaled <- lm(helped_a_stranger ~ I(value / 10000), data = gdp_helped_stranger)
library(stargazer)
stargazer(model_scaled, type = "text",
          title = "Regression: Helped a Stranger vs GDP per Capita (per $10,000)",
          digits = 3,
          covariate.labels = c("GDP per $10,000"),
          dep.var.labels = "Willingness to Help (%)")
```

```{r, warning=FALSE, message=FALSE}
ggplot(gdp_helped_stranger, aes(x = value, y = helped_a_stranger)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x = "GDP per capita (USD)",
    y = "Willingness to help a stranger (%)",
    title = "Relationship between GDP and Willingness to Help a Stranger"
  ) +
  theme_minimal(base_size = 13)
```

To pair visuals with a quantitative reference point, a simple linear regression is included. The goal here is not to establish a causal relationship but to demonstrate how to integrate modeling into a visualization workflow.

In this case, there is a significant negative relationship between GDP per capita and willingness to help a stranger: For every \$10,000 increase in GDP per capita, the percentage of people helping a stranger tends to drop by about 0.8%.

Although the model finds a small but statistically significant negative relationship, GDP explains only a small share (1.6%) of the total variation. This underscores the complexity of prosocial behavior and reinforces the value of visualization for exploring data where simple models may not capture the whole story.

### Conclusion

This walkthrough demonstrates how a mix of R tools—`dplyr` for transformation, `ggplot2` for static summaries - and Plotly for interactive and animated visualizations—can be used to examine global, multi-year datasets. While the helping-stranger question and GDP data serve as the motivating example, the broader goal is to illustrate a reproducible workflow:

-   Creating summary plots to orient the reader

-   Building interactive visuals to uncover spatial and temporal patterns

-   Comparing groups through modular functions

-   Using simple statistical models to complement exploratory graphics

These approaches can be adapted to many datasets and research questions, making them useful for teaching, presentations, dashboards, and exploratory analysis alike.

Feel free to contact me for questions and additional code at ekernan\@hbs.edu.
