---
title: 'Global trends among GDP rankings and the willingness to help strangers'
author: "Liz Kernan, HBS Research Computing Services"
date: "2026-07-16"
categories: [R, data visualization]
execute:
  freeze: true
---

```{r setup, include=FALSE}

#install.packages(c("knitr", "rmarkdown"))
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = TRUE)
```

I am interested in looking at the intersection of social behavior and wealth on a global scale. For this example, I am using world Gallup polling data to the following question: "Have you helped a stranger in the past month?" I then compared these figures to a country's GDP over a ten year span from 2015-2024.

```{r message = FALSE,warning=FALSE}
#load packages 
library(dplyr)
#install.packages("tidyverse")
#install.packages("DT")

library(tidyverse)
library(plotly)
library(crosstalk)
library(DT)
library(ggplot2)
library(gganimate)
library(plotrix)
library(gapminder)
library(viridis)
library(countrycode)
library(ggrepel)
```

```{r, message=FALSE, warning=FALSE}
#load datasets and join

#setwd("/export/projects/rcs_admin_sharedspace/ekernan/happiness_and_taxes/")
countryISO3 <- read.csv("countrycode.csv")
gdp <- read.csv("gdp_per_capita_2015_2024.csv")
helped_stranger <- read.csv("gallup_helped_stranger.csv")
helped_stranger <- helped_stranger[-c(1383:1393), ]

# Convert to ISO3 country codes
helped_stranger$ISO <- countrycode(helped_stranger$country, origin = "country.name", destination = "iso3c")
helped_stranger <- helped_stranger %>% select(-country)

gdp <- gdp %>%
  pivot_longer(
    cols = starts_with("X"),      # pick all columns starting with "X"
    names_to = "year",
    values_to = "value"
  ) %>%
  mutate(year = as.integer(sub("X", "", year)))   # remove "X" and convert to number

gdp_avg <- gdp %>%
  group_by(ISO) %>%
  summarise(avg_gdp = mean(value, na.rm = TRUE))

help_avg <- helped_stranger %>%
  group_by(ISO) %>%
  summarise(avg_help = mean(helped_a_stranger, na.rm = TRUE))

gdp_helped_stranger <- merge(helped_stranger,gdp, by=c("ISO","year"))
gdp_helped_stranger <- merge(gdp_helped_stranger, gdp_avg, by = "ISO")
gdp_helped_stranger <- merge(gdp_helped_stranger, help_avg, by = "ISO")
gdp_helped_stranger$helped_a_stranger<- gdp_helped_stranger$helped_a_stranger * 100
gdp_helped_stranger$avg_help<- gdp_helped_stranger$avg_help * 100

write.csv(gdp_helped_stranger, "pd_data.csv", row.names = FALSE)

avg_help_by_year <- gdp_helped_stranger %>%
  group_by(year) %>%
  summarise(avg_helped = mean(helped_a_stranger, na.rm = TRUE))
```

## Average percentage of willingness to help a stranger by year

```{r, warning=FALSE, message=FALSE, fig.align='center', out.width='70%'}
ggplot(avg_help_by_year, aes(x = factor(year), y = avg_helped,
                             fill = year == 2020)) +   # conditional fill
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "steelblue"), guide = FALSE) +  # red for 2020, blue for others
  geom_text(
    aes(label = paste0(round(avg_helped), "%")),
    vjust = -0.5,
    size = 4
  ) +
  labs(
    title = "Average Helped a Stranger by Year",
    x = "Year",
    y = "Average Helped a Stranger (%)"
  ) +
  theme_minimal() +
  ylim(0, max(avg_help_by_year$avg_helped) * 1.1)
```

This is how respondents answered whether they've helped a stranger in the past month on average from 2015-2024. Notice that there is a big shift between pre and post-covid, as 2020 is highlighted in red.

## Average percentage of willingness to help a stranger among countries over time

```{r, message=FALSE, warning=FALSE}

plot_ly(
  gdp_helped_stranger,
  type = "choropleth",
  locations = ~ISO,
  z = ~helped_a_stranger * 1,
  frame = ~year,
  colorscale = "Reds",
  text = ~paste(
    country),
  colorbar = list(
    title = "% Helped a Stranger",
    tickformat = ".1f"
  )
) %>%
  layout(
    title = "Percentage of willingness to help a stranger per country 2015-2024",
    geo = list(
      projection = list(type = "natural earth"),  # Natural Earth projection
      showframe = TRUE,                           # <-- draws the outer border
      framecolor = "black",                       # <-- color of the outer frame
      framewidth = 2,                             # <-- thickness of the outer frame
      showcoastlines = TRUE,
      coastlinecolor = "gray50",
      showocean = TRUE,                           # <-- adds ocean fill
      oceancolor = "lightblue",                   # <-- ocean color
      landcolor = "lightgray",
      showlakes = TRUE,
      lakecolor = "white",
      showcountries = TRUE,
      countrycolor = "gray60"
    )
  ) %>%
  animation_opts(
    frame = 800,
    transition = 0,
    redraw = TRUE
  )

```

This is a global visualization of how each country answered this question on average over the ten year span using a cool (and free) resource called Plotly. If you hover over each country, you should be able to see its affiliated percentage - and the globe spins!

```{r, message=FALSE, warning=FALSE}
## Average GDP's among countries over time 
#plot_ly(
  #gdp_helped_stranger,
  #type = "choropleth",
  #locations = ~ISO,
  #z = ~value,
  #frame = ~year,
  #colorscale = "Reds",
  #text = ~paste(
    #country),
  #colorbar = list(
    #title = "GDP",
    #tickformat = ".1f"
  #)
#) %>%
  #layout(
    #title = "GDP per country 2015-2024",
    #geo = list(
      #projection = list(type = "natural earth"),  # Natural Earth projection
      #showframe = TRUE,                           # <-- draws the outer border
      #framecolor = "black",                       # <-- color of the outer frame
      #framewidth = 2,                             # <-- thickness of the outer frame
      #showcoastlines = TRUE,
      #coastlinecolor = "gray50",
      #showocean = TRUE,                           # <-- adds ocean fill
      #oceancolor = "lightblue",                   # <-- ocean color
      #landcolor = "lightgray",
      #showlakes = TRUE,
      #lakecolor = "white",
      #showcountries = TRUE,
      #countrycolor = "gray60"
    #)
  #) %>%
  #animation_opts(
    #frame = 800,
    #transition = 0,
    #redraw = TRUE
  #)
```

## Comparing highest and lowest GDP's between 2015 and 2024

```{r, message=FALSE, warning=FALSE}
# helper function to produce the plot for any year
plot_top_bottom_by_year <- function(data, year_val) {
  # select year and top/bottom 10
  df_year <- data %>% filter(year == year_val)
  
  df_top10 <- df_year %>%
    arrange(desc(value)) %>%
    slice_head(n = 10) %>%
    mutate(group = "Top 10 GDP")
  
  df_bottom10 <- df_year %>%
    arrange(value) %>%
    slice_head(n = 10) %>%
    mutate(group = "Bottom 10 GDP")
  
  df_compare <- bind_rows(df_top10, df_bottom10) %>%
    # convert help to percent and build formatted label
    mutate(
      help_percent = helped_a_stranger * 1,
      label_text = paste0(country, " (", sprintf("%.2f%%", help_percent), ")")
    )
  
  ggplot(df_compare, aes(x = value, y = help_percent, color = group, label = label_text)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3, max.overlaps = Inf, segment.color = "grey60") +
    scale_color_manual(values = c("Top 10 GDP" = "blue", "Bottom 10 GDP" = "red")) +
    scale_x_continuous(labels = scales::comma) +
    labs(
      x = paste0("GDP per capita (", year_val, ")"),
      y = "% who helped a stranger",
      color = "Group",
      title = paste0("Top 10 vs Bottom 10 GDP Countries in ", year_val)
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "top"
    )
}

# Example usage
plot_top_bottom_by_year(gdp_helped_stranger, 2015)
plot_top_bottom_by_year(gdp_helped_stranger, 2024)
```

There is a clear shift when comparing 2015-2024 in the percentages that have helped strangers. In 2024, countries with the lowest GDP's helped strangers much more on average than the highest GDP's.

## Comparing highest and lowest GDP's during covid

```{r, message=FALSE, warning=FALSE}
plot_top_bottom_by_year(gdp_helped_stranger, 2020)
```

This difference is interesingly the most dramatic during 2020 and the covid-19 pandemic.

```{r, warning=FALSE, message=FALSE}

# prepare per-year top10 & bottom10
gdp_filtered <- gdp_helped_stranger %>%
  group_by(year) %>%
  arrange(year, desc(value)) %>%
  mutate(rank_desc = row_number(),
         n_in_year = n()) %>%
  filter(rank_desc <= 10 | rank_desc > (n_in_year - 10)) %>%
  ungroup() %>%
  mutate(
    color_group = ifelse(rank_desc <= 10, "Top 10 GDP", "Bottom 10 GDP"),
    helped_percent = helped_a_stranger * 1,                           # convert to percent
    label_text = paste0(country, " (", sprintf("%.2f%%", helped_percent), ")")
  )

# Plotly: formatting similar to your ggplot
plot_ly(
  data = gdp_filtered,
  x = ~value,
  y = ~helped_percent,
  frame = ~year,
  text = ~label_text,
  color = ~color_group,
  colors = c("Top 10 GDP" = "blue", "Bottom 10 GDP" = "red"),
  type = "scatter",
  mode = "markers+text",
  textposition = "middle right",        # label beside marker
  marker = list(size = 10, opacity = 0.85),
  textfont = list(size = 10),
  cliponaxis = FALSE                    # prevents Plotly from cropping labels at axis
) %>%
  layout(
    title = "Top 10 (blue) vs Bottom 10 (red) GDP Countries — GDP vs % Helped over Time",
    xaxis = list(
      title = "GDP per capita",
      range = c(0, max(gdp_filtered$value, na.rm = TRUE) * 1.05),
      tickformat = ",",                  # comma formatting for numbers
      automargin = TRUE
    ),
    yaxis = list(
      title = "% who helped a stranger",
      tickformat = ".2f",
      ticksuffix = "%",
      range = c(0, max(gdp_filtered$helped_percent, na.rm = TRUE) * 1.1),
      automargin = TRUE
    ),
    margin = list(l = 140, r = 40, t = 80, b = 60),   # more left margin so labels fit
    legend = list(title = list(text = "Group"))
  ) %>%
  animation_opts(
    frame = 800,
    transition = 0,
    redraw = TRUE,
    easing = "linear"
  ) %>%
  animation_slider(currentvalue = list(prefix = "Year: ")) %>%
  config(displayModeBar = TRUE)
```

This plot prints out the data for the highest and lowest ten GDP's for each year and where they fall on the scale of helping strangers.

## Statistical Analysis

```{r, warning=FALSE, message=FALSE}
model_scaled <- lm(helped_a_stranger ~ I(value / 10000), data = gdp_helped_stranger)
library(stargazer)
stargazer(model_scaled, type = "text",
          title = "Regression: Helped a Stranger vs GDP per Capita (per $10,000)",
          digits = 3,
          covariate.labels = c("GDP per $10,000"),
          dep.var.labels = "Willingness to Help (%)")
```

```{r, warning=FALSE, message=FALSE}
ggplot(gdp_helped_stranger, aes(x = value, y = helped_a_stranger)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x = "GDP per capita (USD)",
    y = "Willingness to help a stranger (%)",
    title = "Relationship between GDP and Willingness to Help a Stranger"
  ) +
  theme_minimal(base_size = 13)
```

There is a significant negative relationship between GDP per capita and willingness to help a stranger:

For every \$10,000 increase in GDP per capita, the percentage of people helping a stranger tends to drop by about 0.8%.

However, GDP explains only about 1.6% of the variation — so while the effect is real, it’s small.
